# Noah McMahon's Research Question:

# Can we predict a given wine's quality given a list of input attributes?

## K-Nearest Neighbor Approach:

For this research question, we will be using k-Nearest Neighbors, which is a classification method that determines the classification of some metric (in this case quality) based upon a given input's k nearest neighbors (by distance).

```{r Installing & Loading the Necessary Packages}

install.packages("e1071") 
install.packages("caTools") 
install.packages("class") 

library(e1071) 
library(caTools) 
library(class)

set.seed(423) # For Consistency

red_wines$quality <- as.factor(red_wines$quality)
white_wines$quality <- as.factor(white_wines$quality)

```

First, we will split the data into a training and test set (80% training set and 20% test set). Additionally, we will scale the data for both the training and test sets as k-Nearest Neighbors relies on the arithmetic distance between inputs, which needs scaled features so all inputs have a chance to influence the outcome.

```{r Data Splitting & Data Scaling}

split_red <- sample.split(red_wines, SplitRatio = 0.8)
split_white <- sample.split(white_wines, SplitRatio = 0.8)
train_red <- subset(red_wines, split_red == TRUE)
train_white <- subset(white_wines, split_white == TRUE)
test_red <- subset(red_wines, split_red == FALSE)
test_white <- subset(white_wines, split_white == FALSE)

train_red_sc <- scale(train_red[, 1:11])
test_red_sc <- scale(test_red[, 1:11])
train_white_sc <- scale(train_white[, 1:11])
test_white_sc <- scale(test_white[, 1:11])

```

For our next step, we will fit our k-Nearest Neighbor models (one for red wines, one for white wines) to our training sets, starting with k = 1. This means that the quality of a predicted input wine will be determined by the quality of its nearest neighbor.

```{r First k-Nearest Neighbor with k = 1}

red_classifier <- knn(train = train_red_sc, 
                      test = test_red_sc, 
                      cl = train_red$quality, 
                      k = 1)

white_classifier <- knn(train = train_white_sc, 
                      test = test_white_sc, 
                      cl = train_white$quality, 
                      k = 1)

```

We will then display two confusion matrices, one for each model, to start to understand the accuracy of this classification method.

```{r Red Wines Confusion Matrix}

cm_red <- table(test_red$quality, red_classifier)
cm_red

```

```{r White Wines Confusion Matrix}

cm_white <- table(test_white$quality, white_classifier)
cm_white

```

We can now find the accuracy of our classification model for different values of k in order to find the most accurate and best model.

```{r Red Wines Finding Best k}

library(ggplot2)

k_values <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 13, 16, 19, 21, 25, 28, 30, 33, 36, 42, 50, 65, 73, 81, 89, 94, 96, 100, 125, 150)

accuracy_values <- sapply(k_values, function(k) {
  red_classifier <- knn(train = train_red_sc, 
                        test = test_red_sc, 
                        cl = train_red$quality, 
                        k = k)
  1 - mean(red_classifier != test_red$quality)
})

accuracy_data <- data.frame(K = k_values, Accuracy = accuracy_values)

ggplot(accuracy_data, aes(x = K, y = Accuracy)) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "Model Accuracy for Different K Values",
       x = "Number of Neighbors (K)",
       y = "Accuracy") +
  theme_minimal()

```

```{r White Wines Finding Best k}

accuracy_values <- sapply(k_values, function(k) {
  white_classifier <- knn(train = train_white_sc, 
                        test = test_white_sc, 
                        cl = train_white$quality, 
                        k = k)
  1 - mean(white_classifier != test_white$quality)
})

accuracy_data <- data.frame(K = k_values, Accuracy = accuracy_values)

ggplot(accuracy_data, aes(x = K, y = Accuracy)) +
  geom_line(color = "lightgrey", size = 1) +
  geom_point(color = "lightgrey", size = 3) +
  labs(title = "Model Accuracy for Different K Values",
       x = "Number of Neighbors (K)",
       y = "Accuracy") +
  theme_minimal()

```

From the graphs above, we see that our best model is k-Nearest Neighbors with k = 1, which might be due to the fact that using more neighbors might create more noise. However, k-Nearest Neighbors overall does not seem to have great classification accuracy, which leads us to our next model, Random Trees.

## Random Forest Approach:

```{r Installing Necessary Packages}

install.packages("randomForest")
library(randomForest)

```

```{r Setting Up Random Forest Models}

red_classifier_rf = randomForest(x = train_red[-12], 
                             y = train_red$quality, 
                             ntree = 500)

white_classifier_rf = randomForest(x = train_white[-12], 
                             y = train_white$quality, 
                             ntree = 500)

```

```{r Prediction & Confusion Matrix: Red}

red_pred = predict(red_classifier_rf, newdata = test_red[-12])
cm = table(test_red[, 12], red_pred)
cm

```

```{r Prediction & Confusion Matrix: White}

white_pred = predict(white_classifier_rf, newdata = test_white[-12])
cm = table(test_white[, 12], white_pred)
cm

```
```{r Accuracy: Red Wines}

red_rf_accuracy <- 1 - mean(red_pred != test_red$quality)
red_rf_accuracy

```

```{r Accuracy: White Wines}

white_rf_accuracy <- 1 - mean(white_pred != test_white$quality)
white_rf_accuracy

```

```{r Variable Important Plot: Red Wines}

varImpPlot(red_classifier_rf)

```

```{r Variable Important Plot: White Wines}

varImpPlot(white_classifier_rf)

```