---
title: "Final Project"
author: "Caroline Huang"
date: "2025-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(corrplot)
library(car)  # For VIF
```

```{r}
red_wines <- read.csv2("/Users/xiyuehuang/Desktop/STAT 423/wine+quality/winequality-red.csv")
white_wines <- read.csv2("/Users/xiyuehuang/Desktop/STAT 423/wine+quality/winequality-white.csv")
```

```{r}
red_wines$quality <- as.numeric(red_wines$quality)
red_wines$fixed.acidity <- as.numeric(red_wines$fixed.acidity)
red_wines$volatile.acidity <- as.numeric(red_wines$volatile.acidity)
red_wines$citric.acid <- as.numeric(red_wines$citric.acid)
red_wines$residual.sugar <- as.numeric(red_wines$residual.sugar)
red_wines$chlorides <- as.numeric(red_wines$chlorides)
red_wines$free.sulfur.dioxide <- as.numeric(red_wines$free.sulfur.dioxide)
red_wines$total.sulfur.dioxide <- as.numeric(red_wines$total.sulfur.dioxide)
red_wines$density <- as.numeric(red_wines$density)
red_wines$pH <- as.numeric(red_wines$pH)
red_wines$sulphates <- as.numeric(red_wines$sulphates)
red_wines$alcohol <- as.numeric(red_wines$alcohol)

white_wines$quality <- as.numeric(white_wines$quality)
white_wines$fixed.acidity <- as.numeric(white_wines$fixed.acidity)
white_wines$volatile.acidity <- as.numeric(white_wines$volatile.acidity)
white_wines$citric.acid <- as.numeric(white_wines$citric.acid)
white_wines$residual.sugar <- as.numeric(white_wines$residual.sugar)
white_wines$chlorides <- as.numeric(white_wines$chlorides)
white_wines$free.sulfur.dioxide <- as.numeric(white_wines$free.sulfur.dioxide)
white_wines$total.sulfur.dioxide <- as.numeric(white_wines$total.sulfur.dioxide)
white_wines$density <- as.numeric(white_wines$density)
white_wines$pH <- as.numeric(white_wines$pH)
white_wines$sulphates <- as.numeric(white_wines$sulphates)
white_wines$alcohol <- as.numeric(white_wines$alcohol)
```


# Red Wine Analysis
```{r}
lm_test_red <- lm(quality ~ ., data = red_wines)
summary(lm_test_red)
```

```{r}
# Compute Cook’s Distance
cooksd <- cooks.distance(lm_test_red)

# Visualize Cook’s Distance
plot(cooksd, type="h", main="Cook’s Distance", ylab="Cook's Distance")

# Residuals vs. Leverage Plot (Identifies outliers in regression)
plot(lm_test_red, which=5)
```
```{r}
# Compute correlation matrix
cor_matrix <- cor(red_wines)
corrplot(cor_matrix, method="color", type="lower")

# Calculate Variance Inflation Factor (VIF)
vif_values <- vif(lm_test_red)

# Print VIF values
print(vif_values)
```

# White Wine Analysis

```{r}
lm_test_white <- lm(quality ~ ., data = white_wines)
summary(lm_test_white)
```

```{r}
# Compute Cook’s Distance
cooksd <- cooks.distance(lm_test_white)

# Identify influential observations (Threshold: 0.1)
influential_points <- which(cooksd > 0.1)

# Visualize Cook’s Distance
plot(cooksd, type="h", main="Cook’s Distance", ylab="Cook's Distance")
abline(h = 0.1, col="red", lty=2)

# Residuals vs. Leverage Plot (Identifies outliers in regression)
plot(lm_test_white, which=5)
```

```{r}
white_wines[influential_points, ]
```
Observation 2782
	•	Extremely high volatile acidity (0.965) -  Most wines range 0.2 - 0.8.
	•	Extremely high residual sugar (65.8 g/L) -  Normal range is below 10 g/L.
	•	Very high density (1.03898) -  Usually ranges 0.990 - 1.005.
Suspicious: This wine has unusually high residual sugar and volatile acidity, making it a likely sweet, spoiled, or artificially manipulated sample.

Observation 4746
	•	Extremely high free sulfur dioxide (289 mg/L) - Normal range is 10 - 50 mg/L.
	•	Extremely high total sulfur dioxide (440 mg/L) - Legal limits for red wine are ~150 mg/L.
	•	Unusually low density (0.99314) - Suggests watered-down wine or an error.
Likely an anomaly: The extreme sulfur levels and density make this observation highly suspicious—it may be a data entry error.

```{r}
# Remove extreme points only if they are highly influential
white_wine2 <- white_wine[-c(2782, 4746), ]

# Fit model without these points
lm_white2 <- lm(quality ~ ., data=white_wine2)

# Compare Adjusted R² and RMSE
summary(lm_white2)
rmse_white2 <- sqrt(mean(residuals(lm_white2)^2))
print(paste("RMSE (After Removing 2782 & 4746):", rmse_white2))
```


Comparing models about some extreme residuals

```{r}
# Check for extreme standardized residuals (|residual| > 3)
residuals_df <- data.frame(index = 1:nrow(white_wine2), residuals = rstandard(lm_white2))

# Extract extreme residuals
extreme_residuals <- residuals_df[abs(residuals_df$residuals) > 3, ]
```

```{r}
# Remove extreme residuals
wine_cleaned <- white_wine2[-extreme_residuals$index, ]

# Fit new model without extreme outliers
lm_cleaned <- lm(quality ~ ., data=wine_cleaned)

# Compute RMSE
rmse_cleaned <- sqrt(mean(residuals(lm_cleaned)^2))
```

```{r}
# Apply log transformation to quality (to reduce skewness)
white_wine2$log_quality <- log(white_wine2$quality + 1)  # Adding 1 to avoid log(0)

# Fit new model with transformed response
lm_transformed <- lm(log_quality ~ . -quality, data=white_wine2)

# Compute RMSE
rmse_transformed <- sqrt(mean(residuals(lm_transformed)^2))
```

```{r}
comparison <- data.frame(
  Model = c("All Data", "Without Outliers", "Log Transformed"),
  Adjusted_R2 = c(summary(lm_white2)$adj.r.squared, summary(lm_cleaned)$adj.r.squared, summary(lm_transformed)$adj.r.squared),
  RMSE = c(rmse_white2, rmse_cleaned, rmse_transformed)
)

print(comparison)
```

So we prefer drop those extreme residuals. So we will use wine_cleaned data and lm_cleaned model.


```{r}
# Compute correlation matrix
cor_matrix <- cor(wine_cleaned)
corrplot(cor_matrix, method="color", type="lower")

# Calculate Variance Inflation Factor (VIF)
vif_values <- vif(lm_cleaned)

# Convert to a dataframe for better visualization
vif_df <- data.frame(Variable = names(vif_values), VIF = vif_values)

# Sort VIF values in descending order for easy analysis
vif_df <- vif_df[order(-vif_df$VIF), ]

# Print in a readable table format
print(vif_df, row.names = FALSE)
```


```{r}
# Drop the 'density' variable
wine_updated <- wine_cleaned %>% select(-density)

# Fit new model without 'density'
lm_new <- lm(quality ~ ., data=wine_updated)

# Summary of the new model
summary(lm_new)

vif_no_density <- vif(lm_new)
print(vif_no_density)
```

```{r}
# Compute Adjusted R² for both models
adj_r2_old <- summary(lm_cleaned)$adj.r.squared
adj_r2_new <- summary(lm_new)$adj.r.squared

# Compute RMSE for both models
rmse_old <- sqrt(mean(residuals(lm_cleaned)^2))
rmse_new <- sqrt(mean(residuals(lm_new)^2))

# Compute AIC and BIC for both models
aic_old <- AIC(lm_cleaned)
aic_new <- AIC(lm_new)

bic_old <- BIC(lm_cleaned)
bic_new <- BIC(lm_new)

# Create a comparison table
comparison_df <- data.frame(
  Model = c("Original Model", "Without Density"),
  Adjusted_R2 = c(adj_r2_old, adj_r2_new),
  RMSE = c(rmse_old, rmse_new),
  AIC = c(aic_old, aic_new),
  BIC = c(bic_old, bic_new)
)

# Print comparison results
print(comparison_df)
```
