---
title: "Question 3 - Final Project"
author: "Vy Nguyen"
date: "2025-03-02"
output: pdf_document
---

```{r Necessary Packages}
library(tidyverse)
library(ggplot2)
library(car)
library(corrplot)
```

```{r Importing Data}
red_wines <- read.csv2("winequality-red.csv")
white_wines <- read.csv2("winequality-white.csv")
```

First Look at the Data:

```{r First Look}
head(red_wines)

head(white_wines)
```

Casting Data into Appropriate Types:

```{r Casting}

red_wines$quality <- as.numeric(red_wines$quality)
red_wines$fixed.acidity <- as.numeric(red_wines$fixed.acidity)
red_wines$volatile.acidity <- as.numeric(red_wines$volatile.acidity)
red_wines$citric.acid <- as.numeric(red_wines$citric.acid)
red_wines$residual.sugar <- as.numeric(red_wines$residual.sugar)
red_wines$chlorides <- as.numeric(red_wines$chlorides)
red_wines$free.sulfur.dioxide <- as.numeric(red_wines$free.sulfur.dioxide)
red_wines$total.sulfur.dioxide <- as.numeric(red_wines$total.sulfur.dioxide)
red_wines$density <- as.numeric(red_wines$density)
red_wines$pH <- as.numeric(red_wines$pH)
red_wines$sulphates <- as.numeric(red_wines$sulphates)
red_wines$alcohol <- as.numeric(red_wines$alcohol)

white_wines$quality <- as.numeric(white_wines$quality)
white_wines$fixed.acidity <- as.numeric(white_wines$fixed.acidity)
white_wines$volatile.acidity <- as.numeric(white_wines$volatile.acidity)
white_wines$citric.acid <- as.numeric(white_wines$citric.acid)
white_wines$residual.sugar <- as.numeric(white_wines$residual.sugar)
white_wines$chlorides <- as.numeric(white_wines$chlorides)
white_wines$free.sulfur.dioxide <- as.numeric(white_wines$free.sulfur.dioxide)
white_wines$total.sulfur.dioxide <- as.numeric(white_wines$total.sulfur.dioxide)
white_wines$density <- as.numeric(white_wines$density)
white_wines$pH <- as.numeric(white_wines$pH)
white_wines$sulphates <- as.numeric(white_wines$sulphates)
white_wines$alcohol <- as.numeric(white_wines$alcohol)

```

Second Look at Data:

```{r Data After Casting}

head(red_wines)

head(white_wines)

```

# White wine

```{r}

white_wines <- white_wines[-c(2782, 4746), ]

cor_matrix <- cor(white_wines)
print(cor_matrix)
```


# white wine Models

```{r}
# CHeck IVF to decide which factor to use
model_full <- lm(quality ~ ., data = white_wines)
#vif(model)
```

```{r}
model_alcohol <- lm(quality ~ fixed.acidity + volatile.acidity + citric.acid + chlorides + pH + free.sulfur.dioxide + sulphates + total.sulfur.dioxide + alcohol + residual.sugar, data = white_wines)
vif(model_alcohol)
```

```{r}
interaction_model <- lm(quality ~ fixed.acidity + volatile.acidity + citric.acid + chlorides + pH + free.sulfur.dioxide + sulphates + total.sulfur.dioxide + alcohol + residual.sugar + residual.sugar*alcohol, data = white_wines)
```

```{r}
model_no_alcohol <- lm(quality ~ fixed.acidity + volatile.acidity + citric.acid + chlorides + pH + free.sulfur.dioxide + sulphates + total.sulfur.dioxide + residual.sugar, data = white_wines)
```

# Calulate F-test: 

```{r}
summary(interaction_model)
```

```{r}
summary(model_alcohol)
```

```{r}
summary(model_no_alcohol)
```

```{r}
AIC(interaction_model, model_alcohol)
```

We favor model that has the interaction term

## Not include interaction : 

```{r}
anova(model_alcohol, model_no_alcohol)
```

### AIC method: 

```{r}
AIC(model_alcohol, model_no_alcohol)
```


## Interaction vs No Alcohol Model: 

```{r}
AIC(interaction_model, model_no_alcohol)
```
### P-value 

H_0: beta_alcohol = 0

F statistic

```{r}
anova(interaction_model, model_no_alcohol)
```

Since the p-value is extremely small 2.2*e-16, which is well below the 0.01 significance level, we can reject the null hypothesis that the coefficient of alcohol is zero. This provides strong evidence that alcohol has a significant effect on white wine quality.

Furthermore, combining the F-test results with the AIC comparison, we find strong support for alcohol being an important factor in influencing white wine quality.

Thus, higher alcohol content will result in higher quality of wine. 

# Red wine

```{r}
red_correlation <- cor(red_wines)
#print(red_correlation )
```


```{r}
# CHeck IVF to decide which factor to use
red_model_full <- lm(quality ~ ., data = red_wines)
vif(red_model_full)
```

```{r}
red_reduced_model <- lm(quality ~ fixed.acidity + volatile.acidity + citric.acid + chlorides + pH + free.sulfur.dioxide + sulphates + total.sulfur.dioxide + residual.sugar + density, data = red_wines)
```

```{r}
summary(red_model_full)
```


```{r}
AIC(red_model_full, red_reduced_model)
```

```{r}
anova(red_model_full, red_reduced_model)
```





















